### 2020-07-12 paper.tex
add text on traces

% graphics
\usepackage{float}
\usepackage{graphicx}

% math and stuff
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
%\newcommand{\disc}{\mathrm{disc}}
%\newcommand{\cont}{\mathrm{cont}}
%\newcommand{\tr}{\mathtt{tr}}
%\newcommand{\model}{\mathcal{P}}
%\newcommand{\proposal}{\mathcal{Q}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

@@

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}
See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.

\begin{theorem}[Involutive MCMC is stationary]
For $p$, $q$, and $f$ satisfying the requirements above, Algorithm~\ref{alg:involutive-mcmc} defines a probability kernel on $X$ that is stationary with respect to the distribution $p$.
\end{theorem}
\begin{proof}
The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
\end{proof}

@@

\subsection{Traces}
In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the name of a `random choice' to its value.
Following standard terminology in probabilistic programming~\citep{?}, we call these dictionaries \emph{traces} because they can be sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
In the probabilistic programming language used in this paper, the user assigns a unique name or \emph{address} to each random choice using a special syntax:

TODO: include example of tilde syntax

TODO: introduce the mathematical notation for a concrete trace (i.e. $\{a \mapsto 1, b \mapsto 2\}$).

\subsection{A probabilistic program encodes a probability distribution on traces}
For probabilistic programs that only sample individual random choices from discrete probability distributions, the meaning of the program can be described by a probability mass function that assigns a probailtiy to every possible trace.

TODO: Give a simple example of a probabilistic program with discrete random chocies and stochastic control flow, and the resulting probabiilty distribution.
TODO: describe the two variants of tilde syntax

Note that traces are always finite, but there can be no upper bound.
For example, for this program, there is a countably infinite number of possible traces:

TODO: show a discrete probabilistic prgram with the table of traces it could sample

@@

TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and show the reference measure on traces and its density.

% step 1: *assert* that the composition of p and then q defines a probabilistic program with ref meas mu and mut. abs. cont. pi(z) [ok] (( do this without ascribing separate semantics to p and q?? ))
% step 2: show detailed balance on the joint space ... [ok]
% step 3: show stationarity on the joint space [ok, easy from detailed balance]
% step 4: show stationarity with respect to space of p.... ((requires ascribing semantics to p, but maybe not q??)
% - this would be straightforward for a product space, but our space is more complicated?
% - need to assert relationship between p and pi, ...

@@

To define model the auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).
To define the auxiliary densities, we use a second probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
If both programs sample some random choice at the same address $a$ and require different measure spaces $(V_a, \Sigma_a, \mu_a)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}

>>>>

% graphics
\usepackage{float}
\usepackage{graphicx}
\usepackage[usenames,dvipsnames,svgnames]{xcolor}

% math and stuff
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}

% use straight double quotes in math mode
\DeclareMathSymbol{\mathdblquotechar}{\mathalpha}{letters}{`"}
\newcommand{\mathdblquote}{\mathtt{\mathdblquotechar}}
\begingroup\lccode`~=`"\lowercase{\endgroup
  \let~\mathdblquote
}
\AtBeginDocument{\mathcode`"="8000 }

% code listings
\usepackage{listings}
\lstset{
    literate={~} {$\color{black} \bm{\sim}$}{1}
}
\usepackage[framemethod=tikz]{mdframed}
\input{genlisting}
%\definecolor{codegray}{gray}{0.96}Z
\definecolor{addresscolor}{RGB}{22, 88, 12}
\definecolor{distcolor}{RGB}{40, 12, 88}
\definecolor{gencode}{RGB}{255,235,150}
%^\definecolor{jlcode}{RGB}{255, 247, 247}
\definecolor{jlcode}{RGB}{255,210,210}
%\definecolor{dslcode}{RGB}{247, 255, 230}
\definecolor{dslcode}{RGB}{220,242,255}
\usepackage{colortbl}

%\newcommand{\disc}{\mathrm{disc}}
%\newcommand{\cont}{\mathrm{cont}}
%\newcommand{\tr}{\mathtt{tr}}
%\newcommand{\model}{\mathcal{P}}
%\newcommand{\proposal}{\mathcal{Q}}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}
\newcommand{\true}{{\mathtt{T}}}
\newcommand{\false}{{\mathtt{F}}}

@@

\begin{algorithm}[h]
\begin{algorithmic}
\Procedure{involutive-mcmc-move}{$p$, $q$, $f$, $x$}
    \State $y \sim q_x(\cdot)$ \Comment{Sample auxiliary variables}
    \State $(x', y') \gets f(x, y)$ \Comment{Apply involution}
    \State $\alpha \gets
        \displaystyle \frac{p(x') q_{x'}(y')}{p(x) q_{x}(y)} \left( \frac{d (\mu \circ f^{-1})}{d \mu} (x, y)\right)$
    % TODO check it
    \State $r \sim \mathrm{Uniform}(0, 1)$
    \State \algorithmicif \, $r \le \alpha$ \algorithmicthen \, \Return $x'$ \algorithmicelse \, \Return $x$ 
\EndProcedure
\end{algorithmic}
\caption{Involutive MCMC}
\label{alg:involutive-mcmc}
\end{algorithm}

See Section~\ref{sec:radon-nikodym-special-case} for a derivation of the Radon-Nikodym derivative $d (\mu \circ f^{-1}) / d \mu$ for a special case, when the involution $f$ can be factored into (i) an involution on a countable partition of $Z$ and (ii) a family of continuously differentiable bijection on Euclidean spaces.

\begin{theorem}[Involutive MCMC is stationary]
Algorithm~\ref{alg:involutive-mcmc} defines a probability kernel $k$ on $X$ that is stationary with respect to the model probability distribution.
That is, $\int_X k_x(B) p(x) d\mu_X(dx) = \int_B p(x) d\mu_X(dx)$ for all $B \in \Sigma_X$.
\end{theorem}
\begin{proof}
The proof is presented in stages in the appendix (see Section~\ref{sec:involution-detailed-balance}, Section~\ref{sec:involution-is-stationary}, and Section~\ref{sec:involutive-mcmc-is-stationary}).
\end{proof}

@@

In our framework, a probabilistic program defines a probability distribution on finite dictionaries that map the identifer for a random choice (called its \emph{address}) to its value.
Following standard terminology~\citep{?}, we call these dictionaries \emph{traces} because they are often sampled by running a probabilistic program, and `tracing' the execution the program by recording the value of eacn random choice encountered.
The next section makes the concept of a trace more concrete using code in a probabilistic programming language.

\subsection{A Probabilistic Programming Language}

The probabilistic programming language used in this paper is based on the dynamic modeling language in the Gen probabilistic programming system~\citep{gen-pldi}.
In this language, user tags each random choice with an explicit address in the code, as part of a \emph{random-choice expression} of the following form:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
{<address>} ~ <distribution>
\end{lstlisting}}
\end{tabular}
\end{center}
The left-hand side of the expression (green) specifies the address of the choice, and the right-hand side specifies the probability distribution to be sampled from.
For example, the following expression samples a non-negative integer from a geometric distribution and the value gets recorded in the trace at the address \texttt{"a"}:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
{"a"} ~ geometric(0.5)
\end{lstlisting}}
\end{tabular}
\end{center}
In addition to being recorded in the trace, the sampled value is returned to the running program.
That is, the random choice expression evaluates to the sampled value.
For example, the expression below evaluates to a an even non-negative integer:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
({"a"} ~ geometric(0.5)) * 2
\end{lstlisting}}
\end{tabular}
\end{center}
Probabilistic programs are typically constructed by combining multiple random choice expressions with regular Julia code (Gen is embedded in the Julia programming language~\citep{julia}).
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
a = ({"a"} ~ geometric(0.5))
{"b"} ~ bernoulli(a / (a + 1))
\end{lstlisting}}
\end{tabular}
\end{center}
This probabilistic program has an unbounded (countably infinite) number of possible traces it could produce.
One possible trace samples the value 1 at $\texttt{"a"}$ and the Boolean value $\false$ at \texttt{"b"}, and we denote this trace by $\{\mathtt{"a"} \mapsto 1, \mathtt{"b"} \mapsto \false\}$.
Note that the variable to which the first choice was assigned happens to be the same as the address.
This is not required, and often the random choice has no variable name in the program, as in the second random choice with address \texttt{"b"}.
But to reduce possibe unecessary typing, the following optional syntactic sugar will simultaneously assign the value to the program variable $\mathtt{a}$ and record it at trace address $\mathtt{a}$.\footnote{
More precisely, the address is the Julia symbol (interned string) \texttt{a}, constructed with Julia syntax \texttt{:a}.}
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
a ~ geometric(0.5)
\end{lstlisting}}
\end{tabular}
\end{center}
Probabilistic programs can make use of familiar control flow constructs, like loops and branching.
Stochastic control flow is also possible:
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
i = 1
while ({i} ~ bernoulli(0.5))
    i = i + 1
end
\end{lstlisting}}
\end{tabular}
\end{center}
Note that for the probabilistic program to be valid, we require that every address is used at most once within every possible execution.

For probabilistic programs that sample from discrete probability distributions like those used above, the meaning of the program is a probability mass function $p$ that assigns a probability to every possible trace.
For example, for the program with the \texttt{while} loop above, the distribution is:
\[
\begin{array}{c|c}
    x & p(x)\\
    \hline
    \{1 \mapsto \false\} & 0.5\\
    \{1 \mapsto \true, 2 \mapsto \false\} & 0.5^2\\
    \{1 \mapsto \true, 2 \mapsto \true, 3 \mapsto \false\} & 0.5^3\\
    .. & ..
\end{array}
\]
We assume that probabilistic programs halt with probability $1$, meaning that the sum of probabilities over all (finite) traces is one.
Note that traces are always finite, but there may be an infinite number of possible traces, as in the previous example.

In order to construct more complex distributions, we wrap probabilistic programs in functions, called
`generative functions' in Gen.
These functions can then be invoked by other functions, including recursively, which allows programs encode probability distributions on more complex and combinatorial objects.
When invoking such a function, we give an \emph{address namespace} under which every random choice made in that function should be recorded.
This helps prevent addresses of choices defined in different functions from colliding.
The syntax for calling a generative function and assigning the address namespace is the same as that of random choice expressions.
For example:
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function g()
  if ({"go"} ~ bernoulli(0.2))
    n1 = ({"L"} ~ g())
    n2 = ({"R"} ~ g())
    return n1 + n2 + 1
  else
    return 1
  end
end
\end{lstlisting}
\end{tabular}
\end{center}
Because this is a recursive function, a the trace is populated with a hierarchy of addresses.
For example, the first random choice in the first recursive call has address \texttt{"L"."go"}.
When this function is invoked it encodes the following probability distribution on traces (quotation marks omitted from addresses to improve readability):
\[
\begin{array}{c|c}
    x & p(x)\\
    \hline
    \left\{\mathtt{go} \mapsto \false\right\} & 0.8\\
& \\
    \left\{\begin{array}{c}
        \mathtt{go} \mapsto \true, \mathtt{L}.\mathtt{go} \mapsto \false, \mathtt{R}.\mathtt{go} \mapsto \false
    \end{array}\right\} & 0.2 \cdot 0.8^2\\
& \\
    \left\{\begin{array}{c}
        \mathtt{go} \mapsto \true,
        \mathtt{L}.\mathtt{go} \mapsto \true,
        \mathtt{R}.\mathtt{go} \mapsto \false,\\
        \mathtt{L}.\mathtt{L}.\mathtt{go} \mapsto \false,
        \mathtt{L}.\mathtt{R}.\mathtt{go} \mapsto \false
    \end{array}\right\}& 0.2^2 \cdot 0.8^3\\
     .. & ..
\end{array}
\]

@@

TODO: Give an example of a probabilistic program in Gen's dynamic modeling language that includes continuous random choices, and show the reference measure on traces and its density.
\noindent
\begin{center}
\begin{tabular}{c}
{\begin{lstlisting}[basicstyle=\small\ttfamily]
@gen function foo(x1, x2, x3)
  a = ({:a} ~ bernoulli(x1))
  b = ({:b} ~ bernoulli(x2))
  c = ({:c} ~ bernoulli(x3))
  {:d} ~ bernoulli((a && b && c) ? 0.9 : 0.1)
  return a && b && c && d
end
\end{lstlisting}}
\end{tabular}
\end{center}

@@

To define the model and auxiliary densitites used in involutive MCMC in Section~\ref{sec:involutive-mcmc} using probabilistic programs, we write one probabilistic program ($P$) to define the model's probability distribution, and another probabilistic program ($Q$) to define the set of auxiliary distributions.
The model reference measure $(X, \Sigma_P, \mu_P)$ and density that were required in Section~\ref{sec:involutive-mcmc} are set to the probabilistic program's reference measure and density $p$ on traces as defined above (so that $X := T$).
To define the auxiliary densities, we use a second probabilistic program $Q$ that is parametrized by (i.e. takes as input) traces of the model program $P$.
The set of possible inputs to $Q$ is $\{x \in T : p(x) > 0\}$, and for each such $x$ the probabilistic program $Q$ defines a density $q_x : T \to [0, \infty)$ with respect to the reference measure on traces.\footnote{
Note that we the same measure space $(T, \Sigma, \mu)$ on traces for both the model and auxiliary probabilistic program.
If both programs sample some random choice at the same address $a$ and require different measure spaces $(V_a, \Sigma_a, \mu_a)$ for the address, this can be handled formally by making the addresses sampled by the two programs disjoint by prepending either the prefix `p.' or `q.' to addresses.
This is only required for book-keeping in the formalism, and these prefixes do not need to be actually added to the addresses used in the programs.}
